Capstone Movielens Report

### 2025-05-09 reports/capstone-movielens-report.draft4.Rmd
Draft report: documenting User+Movie Effect Model Regularization feature (in progress)

::: {.noteblock data-latex=""}
Here we use the function call [union_cv_results](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L432), which is defined in the script [common-helper.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R), to aggregate the "cross-validation" results.
:::
```{r eval=FALSE}
union_cv_results <- function(data_list) {
  out_dat <- data_list[[1]]

@@

The code of the function [calc_user_movie_effect_MSE](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R#L55) defined in the [UM-effect.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R) script to calculate the `MSE` of the `UME Model` for the given `Test Set` is provided below:
```{r eval=FALSE}
calc_user_movie_effect_MSE <- function(test_set, um_effect){
  mse.result <- test_set |>
    left_join(edx.user_effect, by = "userId") |>
    left_join(um_effect, by = "movieId") |>
    mutate(resid = rating - clamp(mu + a + b)) |> 
    pull(resid) |> mse()

@@

The code of the function [calc_user_movie_effect_RMSE](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R#L51) defined in the [UM-effect.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R) script to calculate the `RMSE` of the `UME Model` for the given `Test Set` is provided below:

@@

Below, we provide the most important code of the training our model using the `5-Fold Cross Validation` method:
```{r eval=FALSE}
  cv.UM_effect <- train_user_movie_effect.cv()
```
```{r}
par(cex = 0.7)
hist(cv.UM_effect$b, 30, xlab = TeX(r'[$\hat{beta}_{j}$)]'),
     main = TeX(r'[Histogram of $\hat{beta}_{j}$]'))
```

@@

### Utilizing Penalized least squares
\
[Section _23.6 Penalized least squares_ of the *Course Textbook*](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#penalized-least-squares) explains why and how we should use _Penalized least squares_ to improve our predictions. The author also explains that the general idea of penalized regression is to control the total variability of the movie effects: $\sum_{j=1}^n \beta_j^2$. Specifically, instead of minimizing the least squares equation, we minimize an equation that adds a penalty:

@@

#### Support function
\
We will use the following function to calculate _RMSE_ in this section:
```{r eval=FALSE}
reg_rmse <- function(b){
  test_set |> 
    left_join(data.frame(userId = as.integer(names(a)), a = a), by = "userId") |>
    left_join(data.frame(movieId = as.integer(names(b)), b = b), by = "movieId") |>
    mutate(resid = rating - clamp(mu + a + b)) |> 
    filter(!is.na(resid)) |>
    pull(resid) |> rmse()
}

@@

@@

>>>>

::: {.noteblock data-latex=""}
Here we use the function call [union_cv_results](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R#L432), which is defined in the script [common-helper.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R), to aggregate the `Cross Validation` results.
:::
```{r eval=FALSE}
union_cv_results <- function(data_list) {
  out_dat <- data_list[[1]]

@@

The code of the function [calc_user_movie_effect_MSE](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R#L55) defined in the [UM-effect.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R) script to calculate the `Mean Squared Error (MSE)` of the `UME Model` for the given `Test Set` is provided below:
```{r eval=FALSE}
calc_user_movie_effect_MSE <- function(test_set, um_effect){
  mse.result <- test_set |>
    left_join(edx.user_effect, by = "userId") |>
    left_join(um_effect, by = "movieId") |>
    mutate(resid = rating - clamp(mu + a + b)) |> 
    pull(resid) |> mse()

@@

The code of the function [calc_user_movie_effect_RMSE](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R#L51) defined in the [UM-effect.functions.R](https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R) script to calculate the `Root Mean Squared Error (RMSE)` of the `UME Model` for the given `Test Set` is provided below:

@@

Below, we provide the most important code of the training our model using the `5-Fold Cross Validation` method:
```{r eval=FALSE}
  cv.UM_effect <- train_user_movie_effect.cv()
```
```{r}
str(cv.UM_effect)

par(cex = 0.7)
hist(cv.UM_effect$b, 30, xlab = TeX(r'[$\hat{beta}_{j}$)]'),
     main = TeX(r'[Histogram of $\hat{beta}_{j}$]'))
```

@@

### Regularizing User+Movie Effect Model

#### Utilizing Penalized least squares
\
[Section _23.6 Penalized least squares_ of the *Course Textbook*](https://rafalab.dfci.harvard.edu/dsbook-part-2/highdim/regularization.html#penalized-least-squares) explains why and how we should use _Penalized least squares_ to improve our predictions. The author also explains that the general idea of penalized regression is to control the total variability of the movie effects: $\sum_{j=1}^n \beta_j^2$. Specifically, instead of minimizing the least squares equation, we minimize an equation that adds a penalty:

@@

We will implement the _Regularization_ method on our  models (starting from the current model) in two steps:

  1. **Preconfiguration:** Preliminary determination of the optimal range of `lambda` values for the `5-Fold Cross Validation` samples;
  2. **Fine-tuning:** figuring out the value of `lambda` that minimizes the model's RMSE.

#### Regularization: Common Helper Functions
\

::: {.noteblock data-latex=""}
The functions described below are defined in the [Model Tuning]() section of the [common-helper.functions.R]() script.
:::

##### `tune.model_param` Function
\

```{r eval=FALSE}
}

@@

::: {.noteblock data-latex=""}
!!! Note!
:::

::: {.noteblock data-latex=""}
!!! Note!
:::

::: {.noteblock data-latex=""}
!!! Note!
:::

::: {.noteblock data-latex=""}
!!! Note!
:::

@@

#### UME Model Regularization: Support Function
\

::: {.noteblock data-latex=""}
The [regularize.test_lambda.UM_effect.cv]() function described below are defined in the [Regularization]() section of the [UM-effect.functions.R]() script.
:::

##### `regularize.test_lambda.UM_effect.cv` Function
\

This function calculates the _RMSE_ for the given $\lambda$ parameter value:
```{r eval=FALSE}
regularize.test_lambda.UM_effect.cv <- function(lambda){
  if (is.na(lambda)) {
    stop("Function: regularize.test_lambda.UM_effect.cv
`lambda` is `NA`")
  }
  um_effect <- train_user_movie_effect.cv(lambda)
  calc_user_movie_effect_RMSE.cv(um_effect)
}
```

::: {.noteblock data-latex=""}
Note that we reuse the function [train_user_movie_effect.cv]() calling it from the [regularize.test_lambda.UM_effect.cv](), but now with the $\lambda$ parameter different from the default (zero) value.
:::

### 2025-05-09 reports/capstone-movielens-report.draft4.tex
Draft report: documenting User+Movie Effect Model Regularization feature (in progress)

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-11-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-38-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-43-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-46-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-51-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-53-1.pdf}

@@

\begin{noteblock}
Here we use the function call
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R\#L432}{union\_cv\_results},
which is defined in the script
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R}{common-helper.functions.R},
to aggregate the ``cross-validation'' results.

@@

The code of the function
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R\#L55}{calc\_user\_movie\_effect\_MSE}
defined in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R}{UM-effect.functions.R}
script to calculate the \texttt{MSE} of the \texttt{UME\ Model} for the
given \texttt{Test\ Set} is provided below:

@@

The code of the function
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R\#L51}{calc\_user\_movie\_effect\_RMSE}
defined in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R}{UM-effect.functions.R}
script to calculate the \texttt{RMSE} of the \texttt{UME\ Model} for the
given \texttt{Test\ Set} is provided below:

@@

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-64-1.pdf}

@@

\subsubsection{Utilizing Penalized least
squares}\label{utilizing-penalized-least-squares}

@@

\paragraph{Support function}\label{support-function}

\hfill\break
We will use the following function to calculate \emph{RMSE} in this
section:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{reg\_rmse }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(b)\{}
\NormalTok{  test\_set }\SpecialCharTok{|\textgreater{}} 
    \FunctionTok{left\_join}\NormalTok{(}\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{userId =} \FunctionTok{as.integer}\NormalTok{(}\FunctionTok{names}\NormalTok{(a)), }\AttributeTok{a =}\NormalTok{ a), }\AttributeTok{by =} \StringTok{"userId"}\NormalTok{) }\SpecialCharTok{|\textgreater{}}
    \FunctionTok{left\_join}\NormalTok{(}\FunctionTok{data.frame}\NormalTok{(}\AttributeTok{movieId =} \FunctionTok{as.integer}\NormalTok{(}\FunctionTok{names}\NormalTok{(b)), }\AttributeTok{b =}\NormalTok{ b), }\AttributeTok{by =} \StringTok{"movieId"}\NormalTok{) }\SpecialCharTok{|\textgreater{}}
    \FunctionTok{mutate}\NormalTok{(}\AttributeTok{resid =}\NormalTok{ rating }\SpecialCharTok{{-}} \FunctionTok{clamp}\NormalTok{(mu }\SpecialCharTok{+}\NormalTok{ a }\SpecialCharTok{+}\NormalTok{ b)) }\SpecialCharTok{|\textgreater{}} 
    \FunctionTok{filter}\NormalTok{(}\SpecialCharTok{!}\FunctionTok{is.na}\NormalTok{(resid)) }\SpecialCharTok{|\textgreater{}}
    \FunctionTok{pull}\NormalTok{(resid) }\SpecialCharTok{|\textgreater{}} \FunctionTok{rmse}\NormalTok{()}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-83-1.pdf}

>>>>

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-12-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-39-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-44-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-47-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-52-1.pdf}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-54-1.pdf}

@@

\begin{noteblock}
Here we use the function call
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R\#L432}{union\_cv\_results},
which is defined in the script
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/data.helper.functions.R}{common-helper.functions.R},
to aggregate the \texttt{Cross\ Validation} results.

@@

The code of the function
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R\#L55}{calc\_user\_movie\_effect\_MSE}
defined in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R}{UM-effect.functions.R}
script to calculate the \texttt{Mean\ Squared\ Error\ (MSE)} of the
\texttt{UME\ Model} for the given \texttt{Test\ Set} is provided below:

@@

The code of the function
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R\#L51}{calc\_user\_movie\_effect\_RMSE}
defined in the
\href{https://github.com/AzKurban-edX-DS/Capstone-MovieLens/blob/main/r/src/support-functions/UM-effect.functions.R}{UM-effect.functions.R}
script to calculate the \texttt{Root\ Mean\ Squared\ Error\ (RMSE)} of
the \texttt{UME\ Model} for the given \texttt{Test\ Set} is provided
below:

@@

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{str}\NormalTok{(cv.UM\_effect)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## tibble [10,677 x 3] (S3: tbl_df/tbl/data.frame)
##  $ movieId: int [1:10677] 1 2 3 4 5 6 7 8 9 10 ...
##  $ b      : num [1:10677] 0.335 -0.306 -0.365 -0.598 -0.444 ...
##  $ n      : num [1:10677] 18907 8593 5574 1253 5065 ...
\end{verbatim}

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-65-1.pdf}

@@

\subsubsection{Regularizing User+Movie Effect
Model}\label{regularizing-usermovie-effect-model}

\paragraph{Utilizing Penalized least
squares}\label{utilizing-penalized-least-squares}

@@

We will implement the \emph{Regularization} method on our models
(starting from the current model) in two steps:

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\tightlist
\item
  \textbf{Preconfiguration:} Preliminary determination of the optimal
  range of \texttt{lambda} values for the
  \texttt{5-Fold\ Cross\ Validation} samples;
\item
  \textbf{Fine-tuning:} figuring out the value of \texttt{lambda} that
  minimizes the model's RMSE.
\end{enumerate}

\paragraph{Regularization: Common Helper
Functions}\label{regularization-common-helper-functions}

\begin{noteblock}
The functions described below are defined in the \href{}{Model Tuning}
section of the \href{}{common-helper.functions.R} script.

\subparagraph{\texorpdfstring{\texttt{tune.model\_param}
Function}{tune.model\_param Function}}\label{tune.model_param-function}

\begin{Shaded}
\begin{Highlighting}[]
\ErrorTok{\}}
\end{Highlighting}
\end{Shaded}

\begin{noteblock}
!!! Note!

\begin{noteblock}
!!! Note!

\begin{noteblock}
!!! Note!

\begin{noteblock}
!!! Note!

\paragraph{UME Model Regularization: Support
Function}\label{ume-model-regularization-support-function}

\begin{noteblock}
The \href{}{regularize.test\_lambda.UM\_effect.cv} function described
below are defined in the \href{}{Regularization} section of the
\href{}{UM-effect.functions.R} script.

\subparagraph{\texorpdfstring{\texttt{regularize.test\_lambda.UM\_effect.cv}
Function}{regularize.test\_lambda.UM\_effect.cv Function}}\label{regularize.test_lambda.um_effect.cv-function}

This function calculates the \emph{RMSE} for the given \(\lambda\)
parameter value:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{regularize.test\_lambda.UM\_effect.cv }\OtherTok{\textless{}{-}} \ControlFlowTok{function}\NormalTok{(lambda)\{}
  \ControlFlowTok{if}\NormalTok{ (}\FunctionTok{is.na}\NormalTok{(lambda)) \{}
    \FunctionTok{stop}\NormalTok{(}\StringTok{"Function: regularize.test\_lambda.UM\_effect.cv}
\StringTok{\textasciigrave{}lambda\textasciigrave{} is \textasciigrave{}NA\textasciigrave{}"}\NormalTok{)}
\NormalTok{  \}}
\NormalTok{  um\_effect }\OtherTok{\textless{}{-}} \FunctionTok{train\_user\_movie\_effect.cv}\NormalTok{(lambda)}
  \FunctionTok{calc\_user\_movie\_effect\_RMSE.cv}\NormalTok{(um\_effect)}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\begin{noteblock}
Note that we reuse the function \href{}{train\_user\_movie\_effect.cv}
calling it from the \href{}{regularize.test\_lambda.UM\_effect.cv}, but
now with the \(\lambda\) parameter different from the default (zero)
value.

@@

\includegraphics{capstone-movielens-report.draft4_files/figure-latex/unnamed-chunk-85-1.pdf}
